---
title: "Class 19: Mutation Project"
auuthor: "Kenny (PID: A18544481)"
format: html
---

I downloaded my sequences and loaded them onto my working/project directory. 

I ran a BLASTp search against REFSEQ_Protein filtered to human only: JOB ID: 


job title: healthy WT
RID: V39DACND014
-- Official symbol:TP53
-- Official full name: tumor protein p53
protein coding --> gene type


> Q2. What are the tumor specific mutations in this particular case (e.g. A130V)?

The tumor specific mutations are at: L137V, V147E, S185R, P190Y


```{r}
library (bio3d)

a <- read.fasta("A18544481_mutant_seq.fa")
```

```{r}
attributes(a)
```

```{r}
a$ali
```


We can score conservation per sequence position
```{r}
score <- conserv(a)
mutation.position <- which( score != 1)
mutation.position
```

```{r}
a$ali[, 137]

a$ali[, 147]

a$ali[, 185]

a$ali[, 190]
```

```{r}
paste0(a$ali[1, mutation.position],
       mutation.position,
       a$ali[2, mutation.position])
```




What PFAM domain do these mutations reside in?

We can use HMMER at the EBI (if it works) to find this information:

OR we can use UniProt 

> Q3. Do your mutations cluster to any particular domain and if so give the name and
PFAM id of this domain? Alternately note whether your protein is single domain and provide
it’s PFAM id/accession and name (e.g. PF00613 and PI3Ka).





