---
title: "Class 18: Pertussis Mini Project"
author: "Kenny Dang (PID: A18544481)"
format: gfm
---

## Background
Pertussis (aka whoopingh cough) is a common lung infection caused by the bacteria B. Pertussis. 

This can infect all ages but is most severe for those under 1 year of age. 

The CDC track the number of [reported cases in US](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html) 
We can "scrape" this data with the **datapasta** package.


```{r}
cdc <- 
  data.frame(
                            year = c(1922,1923,1924,1925,1926,1927,
                                     1928,1929,1930,1931,1932,1933,1934,
                                     1935,1936,1937,1938,1939,1940,
                                     1941,1942,1943,1944,1945,1946,
                                     1947,1948,1949,1950,1951,1952,
                                     1953,1954,1955,1956,1957,1958,
                                     1959,1960,1961,1962,1963,1964,
                                     1965,1966,1967,1968,1969,1970,
                                     1971,1972,1973,1974,1975,1976,1977,
                                     1978,1979,1980,1981,1982,1983,
                                     1984,1985,1986,1987,1988,1989,
                                     1990,1991,1992,1993,1994,1995,
                                     1996,1997,1998,1999,2000,2001,
                                     2002,2003,2004,2005,2006,2007,
                                     2008,2009,2010,2011,2012,2013,
                                     2014,2015,2016,2017,2018,2019,2020,
                                     2021,2022,2023,2024,2025),
    cases = c(107473,164191,165418,152003,202210,181411,
                                     161799,197371,166914,172559,215343,
                                     179135,265269,180518,147237,214652,
                                     227319,103188,183866,222202,191383,191890,
                                     109873,133792,109860,156517,74715,
                                     69479,120718,68687,45030,37129,60886,
                                     62786,31732,28295,32148,40005,14809,
                                     11468,17749,17135,13005,6799,7717,9718,
                                     4810,3285,4249,3036,3287,1759,2402,
                                     1738,1010,2177,2063,1623,1730,1248,
                                     1895,2463,2276,3589,4195,2823,3450,4157,
                                     4570,2719,4083,6586,4617,5137,7796,
                                     6564,7405,7298,7867,7580,9771,11647,
                                     25827,25616,15632,10454,13278,16858,
                                     27550,18719,48277,28639,32971,20762,
                                     17972,18975,15609,18617,6124,2116,3044,
                                     7063,22538, 21996)
  )
  
 
                          
```


> Q1. Make a plot of `year` vs. `cases`. With the help of the R "addin" package datapasta assign the CDC pertussis case number data to a data fram called cdc and use ggplot to make a plot of cases numbers over time. 

```{r}
library (ggplot2)

ggplot (cdc) + 
  aes(year, cases) + 
  geom_point() +
  geom_line() +
  labs (title = "Pertussis Cases by Year (1922-2025)", x = "Year", y = "Number of cases")

```



> Q. Add some major milestones including the first wP vaccine roll-out (1946), the switch to the new aP vaccine (1996), the COVID years (2020)

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

Before the introduction of the wP vaccine, pertussis cases were very high (usually greater than 100,000 per year). After the introduction and use of the wP vaccine, pertussis cases dropped dramatically to very low levels. But after the introduction and switch to the aP vaccine, pertussis cases started increasing slowly again but it is still lower than pre-vaccine levels. 



```{r}
library (ggplot2)

ggplot (cdc) + 
  aes(year, cases) + 
  geom_point() + 
  geom_line() +
  labs (title = "Pertussis Cases by Year (1922-2025)", x = "Year", y = "Number of cases") +
geom_vline(xintercept = 1946, col = "blue", lty=2) +
   geom_vline(xintercept = 1996, col = "red", lty=2) +
 geom_vline(xintercept = 2020, col = "gray", lty=2) +
annotate("text", x = 1946, y = max(cdc$cases), label = "wP", color = "blue") +
annotate("text", x = 1996, y = max(cdc$cases), label = "aP", color = "red")
```



## for cases
22538 for year 2024
21996 for year 2025



> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine, pertussis cases started increasing slightly. A possible explanation for this observed trend is that the aP vaccine provides weaker/shorter-lasting immunity compared to the wP vaccine. Due to the immunity fading over time, more people become as risk for infection which is why there is a rise in pertussis cases after 1996. Also people being more hesitant in regards to vaccinations could be a possible reason as to why there is a slow rise in pertussis cases.  



## There were high case numbers in the pre 1940s. Booster shot helps with decreasing the number of vaccine. wP better than aP vaccine. 




**Why is this vaccine-preventable disease on the upswing?** To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.


#CMI PB project
[Computational Models of Immunity](https://www.cmi-pb.org/) -- Pertussis Boost project aims to provide the scientific community with this very information. 

They make their data avilable via JSON format returning API. We can read this in R with the `read_json()` function from the **jsonlite** package:




```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)

head(subject)
```






> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

There are 87 aP infancy vaccinated subjects in the dataset. There are 85 wP infancy vaccinated subjects in the dataset. 


```{r}
table(subject$infancy_vac)
```


> Q5. How many Male and Female subjects/patients are in the dataset?

There are 60 male patients in the dataset. There are 112 female patients in the dataset. 

```{r}
table(subject$biological_sex)
```







> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

In regards to American Indian/Alaska Native: 0 females and 1 male. In regards to Asians: 32 females and 12 males. In regards to Black or African American: 2 females and 3 males. In regards to more than one race: 15 females and 4 males. In regards to native Hawaiian or other pacific islander: 1 female and 1 male. In regards to unknown or not reported: 14 females and 7 males. In regards to white: 48 females and 32 males. 

```{r}
table(subject$race, subject$biological_sex)
```


> Q. In terms of race and gender, is this dataset representative of the US population?

```{r}
table(subject$race, subject$biological_sex)
```


> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

The average age of wP individuals is 28. The average age of aP individuals is 37. They are significantly different since the individuals who got the wP vaccine are generally older and the difference between the mean ages of the two groups is about 9 years which is a large difference. 

```{r}
library (lubridate)

subject$age <- today() - ymd(subject$year_of_birth)
subject$age
```
```{r}
library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )
```

```{r}
# wP
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```


> Q8. Determine the age of all individuals at time of boost?

The average age of all individuals at time of boost is 26. 

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
age_at_boost
summary(age_at_boost)
```



> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

Based on the histograms below, yes the two groups are significantly different. This conclusion is further supported due to the p-value test as the calculated p-value is 2.37e-23 which is less than 0.05 which means that the difference is statistically significant. Individuals who got the aP vaccine are generally younger while individuals who got the wP vaccine are generally older. 

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```
```{r}
# Or use wilcox.test() 
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```

## Joining multiple tables


Let's read some more database tables:

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)

ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)
```



```{r}
head(specimen)
```



```{r}
head(ab_titer)
```

To analyze this data we need to first "join" (merge/link) the different tables so we have all the data in one place not spread across different tables.

We can use the `*_join()` family of functions from **dplyr** to do this 

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

Completed

```{r}
library (dplyr)

meta <- left_join(specimen, subject, by = "subject_id") 
dim(meta)
head(meta)
```


> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

Completed

```{r}
library (dplyr)

abdata <- inner_join(ab_titer, meta)
dim(abdata)
head(abdata)

```


> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

There are 6698 specimens for the IgE isotype. There are 7265 specimens for the IgG isotype. There are 11993 specimens for the IgG1 isotype. There are 12000 specimens for the IgG2 isotype. There are 12000 specimens for the IgG3 isotype. There are 12000 specimens for the IgG4 isotype. 
```{r}
table(abdata$isotype)
```




> Q. What antigens are reported? 

```{r}

table(abdata$antigen)
```


> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

There are four different datasets which includes 2020, 2021, 2022, and 2023. The most recent dataset has the second most rows. This could be due to the scientists having more samples to process and there is more cases to analyze compared to the last two datasets.

```{r}
unique(abdata$dataset)

table(abdata$dataset)
```

## Examine IgG Ab titer levels

Let's focus on the IgG antigen and make a plot of MFI_normalized for all antigens.

```{r}
igg <- abdata |>
  filter(isotype == "IgG")

head(igg)
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```


> Q. Is there a difference for aP vs wP individuals with these values?

Yes there is a difference between the aP and wP individuals. 

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()
```


> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The antigens FHA, FIM2/3, PT, and PRN show a large differences in the level of IgG antibody titers recognizing them over time. The levels increase around the time of boost and then they decrease later. These antigens are what makes up of the pertussis vaccine so antibody levels change in response to vaccination. Meanwhile, DT, TT, and OVA show little change over time and this could be due to either unrelated control antigens or represent immunity from prior vaccinations that stays relatively stable. 


> Q. Is there a temprol response - i.e do values increase or decrease over time?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```


```{r}
igg %>% filter(visit != 8) %>%
ggplot() +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  xlim(0,75) +
  facet_wrap(vars(infancy_vac, visit), nrow=2) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```



> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```



```{r}
filter(igg, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```


> Q16. What do you notice about these two antigens time courses and the PT data in particular?

The PT levels increase over time and it is higher than FIM2/3. It also looks to reach its peak at visit 7 and then it starts decreasing. This trend is similar for wP and aP subjects. The PT antigen shows a dramatic increase in IgG antibody levels after vaccination, especially around visits 4-6. Meanwhile, the FIM2/3 antigen shows lower antibody levels and a more slow change over time. This means that the immune response is specific to pertussis antigens like PT and not to unrelated antigens like FIM2/3. 


> Q17. Do you see any clear difference in aP vs. wP responses?

Yes, there is a clear difference between aP and wP responses. In many of the visits, individuals who got the wP vaccine show higher IgG antibody levels compared to those who got the aP vaccines. This means that the whole-cell pertussis vaccine may produce a stronger or longer-lasting immune response compared to the acellular pertussis vaccine. The distributions for wP are more towards the right side of the graph compared to aP. 

## Focus on "PT" Pertusisis Toxin antigen


```{r}
pt.igg.21 <- igg |> filter (antigen == "PT",
                       dataset  == "2021_dataset")
pt.igg.21
```



```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```


```{r}
abdata.20 <- abdata %>% filter(dataset == "2020_dataset")

abdata.20 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

> Q18.  Does this trend look similar for the 2020 dataset?

The trend in the 2021 dataset looks similar to the 2020 dataset. In both datasets, IgG PT levels increase after the booster and reaches its peak around day 14 before slowly decreasing over time. The wP group (teal) generally shows higher responses than the aP group (red) in both datasets. But the 2020 dataset shows more variation (more extreme values) and includes some longer follow up time points (going up to 400 days). The 2021 dataset appears to have a more stable trajectory across patients. 

```{r}
ggplot(pt.igg.21) +
  aes(planned_day_relative_to_boost, MFI_normalised, col = infancy_vac, group = subject_id) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept =14, lty = 2)
```



## Obtaining CMI-PB RNASeq data


```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
#meta <- inner_join(specimen, subject)
ssrna <- inner_join(rna, meta)
```


> Q19. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}
ggplot(ssrna) +
  aes(visit, tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```


> Q20. What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

The gene expression reaches its maximum level around visit 4. The gene expression level is low in the beginning and then it increases dramatically as shown around visit 4 and then it decreases around visits 5-7. It reaches its peak again around visit 8.


> Q21. Does this pattern in time match the trend of antibody titer data? If not, why not?

No, the pattern does not match the trend of antibody titer data. When looking at the earlier antibody plots (PT and FIM2/3), the antibody titers increase after vaccination and peak around the middle visits (visit 5-7) but after the peak, they slowly decrease over time. When looking at the gene expression plot, the gene expression level reaches its peak around visit 4 but then it drops dramatically afterwards. This shows that gene expression changes quickly and peaks earlier while antibody titers increase more slowly and last longer. This is due to gene expression representing early immune activation immediately after vaccination while antibody titers represent the protein products of immune cells which takes time to build up and remains in circulation longer. 



```{r}
ggplot(ssrna) +
  aes(tpm, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```



```{r}
ssrna %>%  
  filter(visit==4) %>% 
  ggplot() +
    aes(tpm, col=infancy_vac) + geom_density() + 
    geom_rug() 
```




```{r}
sessionInfo()
```

