---
title: "class011"
author: "Kenny Dang (A18544481)" 
format: html
---

#Comparative protein structure analysis with PCA

We start witha database id "1ake_A"

```{r}
library (bio3d)

id <- "1ake_A"
aa <- get.seq(id)
```

```{r}
aa
```

```{r}
blast <- blast.pdb(aa)
```
```{r}
head(blast$hit.tbl)
```

```{r}
hits <- plot(blast)
```

peak at our "top hits"
```{r}
head(hits$pdb.id)
```



Now we can download these "top hits" these will all be ADK structures in the PDB database. 

```{r}
files <- get.pdb(hits$pdb.id, path = "pdbs", split=TRUE, gzip=TRUE)
```

We need one package from BioConductor, To set this up we need to first install a package called **BiocManager** from CRAN. 

Now we can use the `install()` function from this package like this: 
`BiocManager::install("msa)`

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```



Let's have a wee peak at our structures after "fitting" or superposing:

```{r}
library (bio3dview)
view.pdbs(pdbs)
```

```{r}
view.pdbs(pdbs, colorScheme = "residue")
```
We can run functions like `rmsd()`, `rmsf()` and the best `pca()`
```{r}
pc.xray <- pca (pdbs)
plot (pc.xray)
```

```{r}
plot(pc.xray, 1:2)
```




Finally, let's make a wee movie of the major  "motion" or structural difference in the dataset -- we call this a "trajectory"

```{r}
mktrj(pc.xray, file="results.pdb")
```






